Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-22 Dail_Mad_ext.tex
comments and fixes

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation: \hl{[I'm not familiar with bracket notation,
  so these equations are confusing to me.  How widely used is this
  notation in ecology?  Will others likely understand this?]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_t$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_t$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$). In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is product of the birth rate and the emigration rate,
i.e. the recruitment rate. \citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

@@

Here is some BUGS code:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This model is based
on the \citet{ricker:1954} discrete time version of the logistic population
growth model.  We also implemented the \citet{gompertz:1825} density-dependent
model:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
\label{eq:gomp}
\end{equation}
where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
the interpretations of $r$ and $K$ are similar to in the Ricker model, but
$K$ is constrained to be greater than one.

@@

The autoregressive, population growth, Ricker, and Gompertz
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The population growth plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1}) + \text{Poisson}(\iota)
  \label{eq:expimm1}
\end{equation}
or equivalently
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate.  This model is close to
the constant DM model (equation 4), with $\exp(r)$ instead of $\omega$
and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The Ricker and
Gompertz models can be extended to allow for immigration in the same
way.
\begin{comment}
  Similarly, the Ricker + immigration model can be represented as: the
  Gompertz + immigration model as: (13) and the autoregressive +
  immigration model as: (14)
\end{comment}

@@

The zero inflation factor can also affect
recruitment or population dynamics (see below).
We also considered a zero-inflated negative binomial model of initial
abundance, but this model did not perform well in preliminary tests.

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model (equation
4) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

@@

Our second series of simulations changed dynamics to the
Ricker model.  We used an initial abundance of 10, a maximum
growth rate of 0.005, and a detection probability of 0.25, and
simulated low, medium, and high values of equilibrium
abundance, each with 500 simulations (Table 1).  Our third
series of simulations was based on the Ricker + immigration
dynamics model; here we fixed all parameters the same as the
Ricker model (with K = 10) and simulated low, medium, and high
values of immigration rate with 500 simulation each (Table
1).

@@

Ideas:
I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
II.	Model with low p values, or a range of p values
III.	Ability to distinguish correct initial abundance and dynamics models
IV.	When can you detect DD if it exists (power)?
V.	Building on the last two: what if your model set doesn't
include the correct DD model?  When will the wrong DD model be favored
over the wrong DID model?

@@

We applied these models to North American Breeding Bird Survey (BBS)
data from 1966-2010 for two species in the bordering US states
Maryland and Virginia.  For our focal species, we selected ovenbirds
(\textit{Seiurus aurocapilla}), an abundant and widespread forest breeding
migrant with a stable or increasing trend in the region (CITE) and
golden-winged warblers (\textit{Vermivora chrysoptera}), which only breeds in
the western parts of these states and has been declining in the region
(CITE).  MORE INFORMATION ABOUT THE SPECIES?

GENERAL ABOUT BBS DATA.  We only used data marked as acceptable for
use in the annual BBS analysis.  We summed the total birds of each
species seen on each route and year and used the routes (rather than
the individual stops) as our sites.
Strong winds can interfere with point count observers' ability to hear
birds (\citep{simons_etal:2007}); we tested the effects of wind speed
on detection probability.  BBS volunteers record wind conditions at
the beginning and end of each route on the Beaufort Scale
\citep[start and end wind 0-9]{robbins_etal:1986}.
%(start and end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).
When
start or end wind was not recorded we imputed those values with the
start or end mean.  We took the mean of start and end wind for each
route and put this average wind scale value into four categories: 0 $\leq$
wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an
observer ran a route as a predictor variable for detection
probability.

HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each species,
and then a series of Bayesian models.  We started by testing three
models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with population growth dynamics (equation 7)
and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional models
for p: wind, first, and wind + first.   We selected the minimum AIC
model from that set to test eight additional models of dynamics:
constant, autoregressive, Ricker, Gompertz, autoregressive +
immigration, population growth + immigration, Ricker + immigration,
and Gompertz + immigration.  We then tested the effect of average
minimum temperature for June and July on all dynamics parameters from
the minimum AIC dynamics model.  When run in a maximum likelihood
framework, these models require a maximum abundance to integrate over
(Royle 2004, Dail and Madsen 2011); we used 600 for ovenbirds and 350
for GWWA.

We ran the top ranked models from the maximum likelihood analyses in a
Bayesian framework.  We added random observer effects and, where
appropriate, zero-inflated dynamics.  WILL ADD MORE HEREWe used
non-informative priors.  We tested for lack of convergence using X
Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain
we sampled the MCMC for Y iterations, after Z tuning samples.  Latent
inclusion factors for temperature effects?

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson (Table
1A, models A.1 - A.3).  Even compared to the Poisson, there was no
evidence to support a zero-inflation factor for this species.  The
best supported model for p was additive effects of wind speed and
first run (Table 1A, model A.4).  First run and increasing wind speeds
both decreased p.  All dynamics models with immigration were better
supported than models without immigration (Table 1A, models A.8 -
A.16); the best supported of these was the Ricker + Immigration.  The
best supported model of those without immigration was the Gompertz,
and the autoregressive and the constant model had the least support.
There was little support for an effect of minimum June and July
temperature on dynamics parameters of the Ricker + Immigration model
(Table 1A, models A.17 - A.18).

Estimates for the top ranked model for OVEN
(NB[$\Lambda (.) \alpha (.)$]Ricker+Imm.[$r(.)K(.) \iota (.)$]p(wind+1st)) were similar when
run in the Bayesian framework, except that the estimate of r more than
halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for Bayesian model
estimates we present mean and SD]).  When random observer effects were
added, estimates for $\Lambda$ and K increased dramatically (from 31.6 $\pm$ 4.9
to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1, respectively), and
the estimate for r was intermediate (0.018 $\pm$ 0.006).  The estimate of
the intercept for p (on the logit scale) dropped from -1.5 $\pm$ 0.1 to
-2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  [Richard, I ran
the temperature model in the Bayesian framework (w/ and w/o random
effect) too.  Should we include those results?  As earlier, very
little evidence for an effect of temperature on any of the parameters,
although I couldn't get the latent inclusion parameter to work for r.]
Gelman and Rubin diagnostics and visual examination of the ? plots
provided some evidence for a lack of convergence in estimates of the
intercept for p for all Bayesian models ran.

There was slightly more support for the ZIP distribution for initial
abundance of GWWA than for the negative binomial; both were strongly
supported over the Poisson (Table 1B, models B.1 - B.3).  The best
supported model for p was an effect of first run (Table 1B, model
B.4).  The best supported dynamics model was population growth (Table
1B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
(SE).  There was no support for immigration added to the population
growth or autoregressive models (Table 1B, models B.10 and B.13).
However, the Gompertz + immigration and Ricker + immigration models
were supported over the corresponding models without immigration
(Table 1B, models B.11, B.12, B.14, and B.15), despite low estimates
of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$ 0.002,
respectively).  There was considerable support for an effect of
average minimum June and July temperature on r in the population
growth model (Table 1B, model B.17), with a negative slope.

Because of the closeness of rankings for the ZIP and negative binomial
distributions for GWWA, we also ran the subsequent models with the
negative binomial initial abundance.  Rankings of subsequent models
were similar, but with a p covariate (first) the negative binomial
distribution outranked the ZIP.

STILL WORKING ON JAGS GWWA RESULTS
Estimated detection probabilities were generally low.  For OVEN, the
estimated probability of detecting a bird with no wind and not the
first time an observer had run a route varied between 0.046 $\pm$ 0.002
(model A.16) and 0.192 $\pm$ 0.015 (model A.11), among models that
accounted for both.  For GWWA, the estimated probability of detecting
a bird by a non-first time observer varied between 0.025 $\pm$ 0.008
(model B.18) and 0.141 $\pm$ 0.027 (model B.16), among models that
accounted for first but not wind.

@@

A unique aspect of the Dail and Madsen model as originally devised was
that it allowed for the estimation of demographic parameters under
strict distributional assumptions and with the assumption of
geographic closure. Clearly, estimating demographic parameters from
count data is a loftyn ambitious goal, and the required assumption
will not be valid in many cases. Nonetheless, their approach is
important because it allows for the combination of count data with
other data types such as capture-recapture data , which offers a
means of making inferences about demographic processes at broad
spatial scales. Models combining count data and demographic data are
often referred to as integrated population models, and This topic has
received much attention lately under t

@@

[paragraph on zero-inflation]The zero-inflated models for initial
abundance and dynamics allow one to estimate and model the proportion
of sites outside of the range of a species.  This could be especially
useful when different factors control the range than the abundance or
dynamics within the range.  [some examples where that is the case?]

>>>>

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation: \hl{[I'm not familiar with bracket notation,
  so these equations are confusing to me.  How widely used is this
  notation in ecology?  Will others likely understand this?  Also, we might want
to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_t$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_t$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$). In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum\hl{[I think!]} of the birth rate and the emigration rate,
i.e. the recruitment rate. \hl{[Feels a little repetitious with the material 
before the equation, although I understand you're describing S and G 
there and omega and gamma here.  In general, these sections 
describing the DM models are clear and well written, but we could 
possibly spend less time on them b/c it's already published.  
Possibly it's better this way, though.]} \citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

@@

Here is some BUGS code: \hl{[This is code for a model we haven't
presented yet.  Could we move these sections after model extensions?]}
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.~\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This is a stochastic version
of the Ricker-logistic population growth model \citep{ricker:1954}.  We also 
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1)))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the Ricker model.

@@

The autoregressive, population growth, Ricker, and Gompertz
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The population growth plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is constrained to be 
positive (this is equivalent to separate Poisson processes for growth and 
immigration).  This model is close to
the constant DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead of $\omega$
and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The Ricker and
Gompertz models can be extended to allow for immigration in the same
way.
\begin{comment}
  Similarly, the Ricker + immigration model can be represented as: the
  Gompertz + immigration model as: (13) and the autoregressive +
  immigration model as: (14)
\end{comment}

@@

%The zero inflation factor can also affect
%recruitment or population dynamics (see below).
We also considered a zero-inflated negative binomial model of initial
abundance, but this model did not perform well in preliminary tests.

@@

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model 
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{it} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\mathrm{Poisson}(\gamma) &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textrm{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

@@

Our second series of simulations changed dynamics to the Ricker model.  
We used an initial abundance of 10 and a detection probability of 0.25, 
and simulated low, medium, and high values of equilibrium abundance 
and maximum growth rate, each with 1000 simulations (Table 1; 
9 total combinations).  Our third series of simulations was based on the 
Ricker + immigration dynamics model; here we fixed all parameters the 
same as the Ricker model (with r = 0.05 and K = 10) and simulated 
low, medium, and high values of immigration rate with 1000 simulation 
each (Table 1).

@@

\begin{comment}
Ideas:
I.	Test importance of random effects modeling, and other differences between JAGS and unmarked
II.	Model with low p values, or a range of p values
III.	Ability to distinguish correct initial abundance and dynamics models
IV.	When can you detect DD if it exists (power)?
V.	Building on the last two: what if your model set doesn't
include the correct DD model?  When will the wrong DD model be favored
over the wrong DID model?
\end{comment}

@@

We applied these models to North American Breeding Bird Survey (BBS)
data from 1966-2010 for two species in the bordering US states
Maryland and Virginia.  For our focal species, we selected ovenbirds
(\textit{Seiurus aurocapilla}), an abundant and widespread forest breeding
migrant with a stable or increasing trend in the region (CITE) and
golden-winged warblers (\textit{Vermivora chrysoptera}), which only breeds in
the western parts of these states and has been declining in the region
(CITE).  %MORE INFORMATION ABOUT THE SPECIES?

%GENERAL ABOUT BBS DATA.  
The BBS is an annual roadside survey implemented by trained observers 
in the United States and Canada. An observer conducts 50 3-minute point counts, 
with 400 m radiuses and 0.8 km apart from each other along a 39.4 km route. 
We only used data marked as acceptable for use in the annual BBS analysis 
\citep{sauer_etal:1994auk}.  We summed the total birds of each
species seen on each route and year and used the routes (rather than
the individual stops) as our sites.
Strong winds can interfere with point count observers' ability to hear
birds \citep{simons_etal:2007}; we tested the effects of wind speed
on detection probability.  BBS volunteers record wind conditions at
the beginning and end of each route on the Beaufort Scale
\citep[start and end wind 0-9]{robbins_etal:1986}.
%(start and end wind; Robbins et al. 1986)(start and end wind; 0-9; cite).
When
start or end wind was not recorded we imputed those values with the
start or end mean.  We took the mean of start and end wind for each
route and put this average wind scale value into four categories: 0 $\leq$
wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind $\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an
observer ran a route as a predictor variable for detection
probability.

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each species,
and then a series of Bayesian models.  We started by testing three
models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with population growth dynamics (Eq.~\ref{eq:exp})
and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional models
for p: wind, first, and wind + first.   We selected the minimum AIC
model from that set to test eight additional models of dynamics:
constant, autoregressive, Ricker, Gompertz, autoregressive +
immigration, population growth + immigration, Ricker + immigration,
and Gompertz + immigration.  %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.  
As above, these models require a maximum abundance to integrate over
when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350
for GWWA.

We ran the top ranked models from the maximum likelihood analyses in a
Bayesian framework.  We added random observer effects and, where
appropriate, zero-inflated dynamics.  We used
non-informative priors.  We tested for lack of convergence using 5
Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain
we sampled the MCMC for at least 40,000 iterations, after at least 3,000 tuning samples.  
%Latent inclusion factors for temperature effects?

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson (Table
2A, models A.1 - A.3).  Even compared to the Poisson, there was no
evidence to support a zero-inflation factor for this species.  The
best supported model for p was additive effects of wind speed and
first run (Table 2A, model A.4).  First run and increasing wind speeds
both decreased p.  All dynamics models with immigration were better
supported than models without immigration (Table 2A, models A.8 -
A.16); the best supported of these was the Ricker + Immigration.  The
best supported model of those without immigration was the Gompertz,
and the autoregressive and the constant model had the least support.
%There was little support for an effect of minimum June and July
% temperature on dynamics parameters of the Ricker + Immigration model
%(Table 2A, models A.17 - A.18).

Estimates for the top ranked model for OVEN
(NB[$\Lambda (.) \alpha (.)$]Ricker+Imm.[$r(.)K(.) \iota (.)$]p(wind+1st)) were similar when
run in the Bayesian framework, except that the estimate of r more than
halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for Bayesian model
estimates we present mean and SD]).  When random observer effects were
added, estimates for $\Lambda$ and K increased dramatically (from 31.6 $\pm$ 4.9
to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1, respectively), and
the estimate for r was intermediate (0.018 $\pm$ 0.006).  The estimate of
the intercept for p (on the logit scale) dropped from -1.5 $\pm$ 0.1 to
-2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
%[Richard, I ran
%the temperature model in the Bayesian framework (w/ and w/o random
%effect) too.  Should we include those results?  As earlier, very
%little evidence for an effect of temperature on any of the parameters,
%although I couldn't get the latent inclusion parameter to work for r.]
Gelman and Rubin diagnostics and visual examination of the ? plots
provided some evidence for a lack of convergence in estimates of the
intercept for p for all Bayesian models ran.

There was slightly more support for the ZIP distribution for initial
abundance of GWWA than for the negative binomial; both were strongly
supported over the Poisson (Table 2B, models B.1 - B.3).  The best
supported model for p was an effect of first run (Table 2B, model
B.4).  The best supported dynamics model was population growth (Table
2B, model B.8); the estimate of r from this model was -0.058 $\pm$ 0.017
(SE).  There was no support for immigration added to the population
growth or autoregressive models (Table 1B, models B.10 and B.13).
However, the Gompertz + immigration and Ricker + immigration models
were supported over the corresponding models without immigration
(Table 1B, models B.11, B.12, B.14, and B.15), despite low estimates
of $\iota$ from these models (0.003 $\pm$ 0.003 and 0.003 $\pm$ 0.002,
respectively).  %There was considerable support for an effect of
%average minimum June and July temperature on r in the population
%growth model (Table 1B, model B.17), with a negative slope.
Because of the closeness of rankings for the ZIP and negative binomial
distributions for GWWA, we also ran the subsequent models with the
negative binomial initial abundance.  Rankings of subsequent models
were similar, but with a p covariate (first) the negative binomial
distribution outranked the ZIP.

Estimates for the top ranked model for GWWA (ZIP[Λ(.)α(.)]Trend[r(.)]p(1st)) 
were similar when run in the Bayesian framework.  When random observer 
effects were added, the estimate for Λ  increased dramatically 
(from 22.4 ± 7.6 to 55.5 ± 30.9).  The estimate of the intercept for p (on the 
logit scale) dropped from -3.5 ± 0.2 to -5.6 ± 1.1, and the estimate of σp 
was 1.7 ± 0.7.  Adding zero-inflation to the dynamics (as well as the initial 
abundance) has no effect for trend models, because populations that started 
at 0 would stay at 0 in any case.  Therefore, we tested zero-inflated dynamics 
on the Trend + Imm. model.  Adding zero-inflation to the dynamics increased 
the estimate of ι from 0 to 0.091 ± 0.049, but had only minor effects on the 
other parameters.  Gelman and Rubin diagnostics and visual examination of 
the ? plots provided evidence for a lack of convergence in estimates of Λ 
for all Bayesian models ran, and for most parameters when random effects 
were included.

Estimated detection probabilities were generally low.  The estimated probability 
of detecting an OVEN in the top model (A.8) varied between 0.157 ± 0.014 
and 0.184 ± 0.015, depending on wind speed and whether the observer had 
run that route before.  The estimated probability of detecting a GWWA went 
from 0.013 ± 0.006 for first time observers to 0.028 ± 0.008 for repeat 
observers for the top model (B.8).  Accounting for random observer effects 
decreased average detection probability for both species.

@@

A unique aspect of the Dail and Madsen model as originally
devised was that it allowed for the estimation of demographic
parameters under strict distributional assumptions and with the
assumption of geographic closure. Clearly, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases. One
advantage of the original DM models over our extensions here is
that they model population dynamics more mechanistically, by
focusing on survival and recruitment. As we report here, these
models often provide unrealistic estimates of these vital rates
and are generally less supported in model selection. Adding
density dependence directly to one or both vital rates may solve
both of these problems, although this might require very good
data. Another strategy would be to use estimates of vital rates
from demographic studies as priors in a Bayesian analysis using
DM models. Finally, count data could be combined with other data
types such as capture-recapture data, which offers a means of
making inferences about demographic processes at broad spatial
scales. Models combining count data and demographic data are
often referred to as integrated population models (IPM; CITE),
and we believe that a DM model would provide an effective part
of an IPM.

@@

The zero-inflated models for initial
abundance and dynamics allow one to estimate and model the proportion
of sites outside of the range of a species.  This could be especially
useful when different factors control the range than the abundance or
dynamics within the range.  [some examples where that is the case?]
